---
title: "Introduction"
subtitle: "Urban Economics"
author: "Dr. Hussain Hadah"
date: "Spring 2026"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    chakra: "libs/remark-latest.min.js"
    css: ["default", "css/ath-slides.css", "css/ath-inferno-fonts.css", "css/animate.css"]
    seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
      # beforeInit: "libs/offline-mathjax.js"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```

```{r packages-data, include=FALSE}
library(tidyverse)
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "share_again"))
```

class: center, middle, inverse, title-slide

# ECON 414 - Urban Economics
## Introduction
### Dr. Hussain Hadah
### Spring 2026

---











# What is Urban Economics? 



* Urban Economics intersects economics and geography studying the location choices of economic activity 
 - Identifies economic forces shaping the urban structure within/across cities
 - What are the costs and benefits of cities? 
 - What public policies promote efficiency in cities?

--

* Why location matter?
 - Agglomeration forces
 - Externalities, neighborhood effects

--

* What determines people's/firms' willingness to pay for places? 
 - Why do firms stay in places where they pay high salaries?



---

class: inverse,center, middle

# Regularities Across cities


---
# Population Density in the US



<iframe src="maps/popdens.html"  style="width: 1200px; height: 500px; border: 5px" alt="">


---

# Household Income in the US


<iframe src="maps/hhincome.html"  style="width: 1200px; height: 500px; border: 5px" alt="">

 


---
class: inverse, middle, center
# Regularities Within Cities


---
# Population Density in Chicago-IL

<iframe src="maps/popdens_chi.html"  style="width: 1200px; height: 500px; border: 5px" alt="">
---
# Commuting in Chicago-IL

<iframe src="maps/commuting_chi.html"  style="width: 1200px; height: 500px; border: 5px" alt="">
---

class: inverse, middle, center

# What are cities? What's good about it?

---

# What are cities? What's good?


* A city is a relatively small area that contains a large number of people

 - Three conditions must be satisfied for a city to form: Agricultural surplus, urban production, transportation system
  
 - The share of people living in cities in the US increased from 2% to 82% between 1700 and 2010. Technology advances caused this movement. Why?
 
 - **Cities increase our standard of living**

--

*  Thicker markets offer better matches between workers, firms, and consumers
 
 - Productivity benefits 
 
 - Transportation costs 
 
 - Knowledge spillovers
 
 - Consumer cities
 
 
---
# Urbanization in the US

<iframe src="https://ourworldindata.org/grapher/urban-and-rural-population-stacked?country=~USA" style="width: 1200px; height: 500px; border: 5px" alt="">

---
# Urbanization across the World

<iframe src="https://ourworldindata.org/grapher/share-of-population-urban?year=1960" style="width: 1200px; height: 500px; border: 5px" alt="">
---
# Share of people living in slums
<iframe src="https://ourworldindata.org/grapher/share-of-urban-population-living-in-slums?tab=chart&time=earliest&country=IND~TUR~BRA~BGD~ZAF~NGA~CHN&region=Africa" style="width: 1200px; height: 500px; border: 5px" alt="">

---
class: inverse, middle, center

# The cost of cities

---

# COVID-19 Distribution across US 04/22/2020


<iframe src="maps/covid.html"  style="width: 1200px; height: 500px; border: 5px" alt="">


---
# COVID-19 Distribution across US 01/24/2021



<iframe src="maps/covid2.html"  style="width: 1200px; height: 500px; border: 5px" alt="">


---

class: inverse, middle, center
# The Economic approach to cities


---

# The Economic approach to cities


- Focus on people rather than places

--

- Three key elements:
  - People respond to incentives 
  - There is no free lunch (no-arbitrage equilibrium) 
  - Good policies increase people's range of choices 

--

- Relies on a spatial equilibrium for workers, employers, and developers
  - Individuals are indifferent across space: 
  
  Income + Amenities - Housing Costs - Transportation Costs **is constant over space**
  - Differences in wages must be offset by differences in productivity
  - Developers are indifferent about building or not new units

---

# Axioms of Urban Economics

1. Prices adjust to achieve locational equilibrium
  - Prices will adjust, so no one has the incentive to move

--
  
2. Self-reinforcing effects generate extreme outcomes
  - A change in something leads to additional changes in the same direction

--

3. Externalities cause inefficiencies
  - A crowded highway, air pollution, etc.

--

4. Production is subject to Economies of Scale
  - The average cost of production decreases as output increases

--

5. Competition generates zero economic profit
  - Zero economic profit means earning "normal" accounting profit
